---
output: html_document
---

# Deconvolution script

An example signature matrix from CIBERSORTX LM22 looks like this:

```{r ,warning=FALSE}
library(knitr)
setwd("~/Desktop/PhD/2024 Second Term/deconvolution")

LM22<- read.delim("LM22.txt")
kable(head(LM22[,1:5]))
```

preparing signature matrix from celina's analysis for synovium and loading raw counts

```{r}
raw_counts_from_scRNA <- read.csv("from celina new/rstudio-export/gene_raw_counts_per_cell_type.csv") 
cell_markers_all <- read.csv("from celina new/Cell_type_markers.csv")
kable(head(cell_markers_all))
unique(cell_markers_all$annotation)

#we will pick top 25 and top 50 genes for each cluster
#only T cells, Fibroblasts, Endothelial, B cells and Myeloids will be taken into account
cell_markers_top50 <- data.frame(gene = character(), annotation=character())
cell_markers_top25 <- data.frame(gene = character(), annotation=character())
cell_type_vector <- c("Myeloid cells", "Fibroblasts", "T cells", "Endothelial cells", "B cells")

for (cell_type in cell_type_vector) {
  cell_markers_top50 <- rbind(cell_markers_top50, cell_markers_all[cell_markers_all$annotation == cell_type, c("gene","annotation") ][1:50,])
  cell_markers_top25 <- rbind(cell_markers_top25, cell_markers_all[cell_markers_all$annotation == cell_type, c("gene","annotation") ][1:25,] )
}
unique(cell_markers_top50$annotation)

```

now we have to check if the genes representing the cell types are unique to cell types

```{r}
#for top 25
all_unique <- length(cell_markers_top25$gene) == length(unique(cell_markers_top25$gene))
all_unique
duplicates <- cell_markers_top25$gene[duplicated(cell_markers_top25$gene)]
duplicates
# which cell type has this marker gene more? 
for (dup in duplicates){
  kable(cell_markers_all[cell_markers_all$gene == dup,])
}

#more in T cells so we delete it from the B cells
cell_markers_top25 <- cell_markers_top25[! (cell_markers_top25$gene == duplicates[1] & cell_markers_top25$annotation =="B cells"),] 

#for top 50
all_unique <- length(cell_markers_top50$gene) == length(unique(cell_markers_top50$gene))
print(all_unique)
duplicates <- cell_markers_top50$gene[duplicated(cell_markers_top50$gene)]
print(duplicates)

#manually checked and all of them are more in T cells compared to B cells
for (dup in duplicates) {
  cell_markers_top50 <- cell_markers_top50[!(cell_markers_top50$gene == dup & cell_markers_top50$annotation =="B cells"),] 

}

```

raw counts are 'processed' so we have to turn them back to integers

```{r}
raw_counts_from_scRNA[, -1] <- lapply(raw_counts_from_scRNA[, -1], function(x) round(x))
kable(head(raw_counts_from_scRNA[,1:5])) 
```

merge the cell marker genes with the raw counts

```{r}
#signature matrix for top 25
signature_matrix_top25 <- raw_counts_from_scRNA[raw_counts_from_scRNA$Gene %in% cell_markers_top25$gene,]
colnames(signature_matrix_top25)
signature_matrix_top25 <- signature_matrix_top25[,c("Gene","B.cells","Endothelial.cells","Fibroblasts","Myeloid.cells","T.cells")]
nrow(signature_matrix_top25)
nrow(cell_markers_top25)
#none of the genes are lost :)

#signature matrix for top 50
signature_matrix_top50 <- raw_counts_from_scRNA[raw_counts_from_scRNA$Gene %in% cell_markers_top50$gene,]
signature_matrix_top50 <- signature_matrix_top50[,c("Gene","B.cells","Endothelial.cells","Fibroblasts","Myeloid.cells","T.cells")]

nrow(signature_matrix_top50)
nrow(cell_markers_top50)

```

we do CPM normalization 

```{r}
cpm_normalization <- function(df) { 
  total_counts <- colSums(df[,-1]) # Total counts for each column 
  cpm_normalized <- t(t(df[,-1]) /total_counts * 1e6) # Normalize and scale 
  df[,-1] <- cpm_normalized
  return (df) 
  }

signature_matrix_top25 <- cpm_normalization(signature_matrix_top25)
signature_matrix_top50 <- cpm_normalization(signature_matrix_top50)

```
read bulk rna raw data 

```{r}
# count matrix of GSE89408
 #this doesnt have the raw counts for some reason, we will only get the column names
count_matrix <- read.delim("bulk_raw_data/GSE89408_GEO_count_matrix_rename.txt") 
column_names_to_append <- colnames(count_matrix[2:ncol(count_matrix)])

#create a list 
gsm_ids <- paste0("GSM", 2370970:2371187)
gsm_list <- setNames(as.list(column_names_to_append), gsm_ids) #kind of a dictionary in python

#this one has the raw data (from new feature of GEO)
count_matrix <- read.delim("bulk_raw_data/GSE89408_raw_counts_GRCh38.p13_NCBI.tsv")
old_colnames <- colnames(count_matrix)
new_colnames <- old_colnames  # Copy to keep structure

#update the column names
for (i in 2:length(old_colnames)) {
  if (old_colnames[i] %in% names(gsm_list)) {
    new_colnames[i] <- gsm_list[[old_colnames[i]]]
  }
}
colnames(count_matrix) <- new_colnames

#change from GeneID to hgnc symbols
geneids <- read.delim("bulk_raw_data/Human.GRCh38.p13.annot 2.tsv")
geneids <- geneids[c(1,2)]

count_matrix <- merge(geneids, count_matrix, by = "GeneID", all.y = TRUE)

all_unique <- length(count_matrix$Symbol) == length(unique(count_matrix$Symbol))
print(all_unique)

duplicates <- count_matrix$Symbol[duplicated(count_matrix$Symbol)]

print(duplicates)
```
for some reason trnav-cac is represented 3 times 

```{r}
kable(count_matrix[count_matrix$Symbol == "TRNAV-CAC",])

#I will get rid of the first and third one -> less reads mapped 

count_matrix <- count_matrix[-c(37317,37405),]
all_unique <- length(count_matrix$Symbol) == length(unique(count_matrix$Symbol))
print(all_unique)

count_matrix$GeneID <- NULL
colnames(count_matrix)[colnames(count_matrix) == "Symbol"] <- "Gene"
count_matrix_normalized <- cpm_normalization(count_matrix)
#now it is ready
```
now that count matrix is ready we make the signature matrix genes comparable to this

update gene names (everything to hgnc symbol)

```{r, warning =FALSE}
library(biomaRt)
library(dplyr)

# Connect to the Ensembl database
ensembl <- useMart("ensembl", dataset = "hsapiens_gene_ensembl")

# Retrieve updated gene names
gene_synonyms <- getBM(
  attributes = c("hgnc_symbol", "external_synonym"),
  filters = "external_synonym",
  values = count_matrix$Gene,
  mart = ensembl
) #this gets all synonyms

#some hgnc symbols are lacking so get rid of them
#this is VERY cruical: some synoyms map to more than one hgnc symbols, almost always the first one is the true one
gene_synonyms_updated <- gene_synonyms[gene_synonyms$hgnc_symbol !="",]
gene_synonyms_updated <- gene_synonyms_updated %>%
  distinct(external_synonym, .keep_all = TRUE)
gene_synonyms_updated <- gene_synonyms_updated %>%
  distinct(hgnc_symbol, .keep_all = TRUE)
```

update the names of the genes in signature matrixes

```{r}
update_name <- function(df){
  
  gene_synonyms_updated_unique <- gene_synonyms_updated[!(gene_synonyms_updated$hgnc_symbol %in% df$Gene),]
  
  df <- merge(
    df, 
    gene_synonyms_updated_unique, 
    by.x = "Gene", 
    by.y = "external_synonym",  # Matching based on synonym
    all.x = TRUE                  # Keep all rows from count_matrix
  )
  
  # update the gene names
  df$Gene <- ifelse(
    !is.na(df$hgnc_symbol),  # if hgnc symbol is available
    df$hgnc_symbol,         # replace 
    df$Gene                 # keep the original name 
  )
  
  df$hgnc_symbol <- NULL

  return (df)
}

signature_matrix_top25 <- update_name(signature_matrix_top25)
signature_matrix_top50 <- update_name(signature_matrix_top50)

```

we *can* now check if every gene is in count matrix (of course not)

```{r}
df_merged <- merge(signature_matrix_top50, count_matrix_normalized, by = "Gene", all.x = TRUE) 

na_rows <- df_merged[rowSums(is.na(df_merged)) > 0, ]
kable(na_rows[,1], caption = "genes not present in count data")
```

these were manually checked and only one of them can be modified, other than that nothing left to do here

```{r}
signature_matrix_top50[signature_matrix_top50$Gene == "PALM2-AKAP2",]$Gene <- "PALM2AKAP2"

```


now we can save count_matrix and signature_matrixes for cibersortx

```{r}
#write.table(signature_matrix_top25, "for_input/signature_matrix_top25_synovium.txt", sep ="\t", row.names = FALSE, quote = FALSE) 
#write.table(signature_matrix_top50, "for_input/signature_matrix_top50_synovium.txt", sep ="\t", row.names = FALSE, quote = FALSE) 
#write.table(count_matrix_normalized,"for_input/count_matrix_GSE89408_normalized.txt", sep = "\t", row.names = FALSE, quote = FALSE)
#write.table(count_matrix,"for_input/count_matrix_GSE89408.txt", sep = "\t", row.names = FALSE, quote = FALSE)

```
