#Visualization for Deconvolution

From the CIBERSORTX deconvolution we visualize the cell proportions
```{r,echo=FALSE,warning=FALSE}
library(tidyr)
library(ggplot2)
library(RColorBrewer)
library(tidyverse)
library(rmarkdown)
library(knitr)
library(ggh4x)

```
First organize the data
```{r}
setwd("~/Desktop/PhD/2024 Second Term/deconvolution")
data <- read.delim("results/CIBERSORTx_signature_matrix_top50.txt")
colnames(data)
colnames(data) <- c("Mixture","B cells","Endothelial Cells","Fibroblasts","Myeloid Cells","T cells","pvalue","correlation","rmse")

data <- data %>%
  mutate(Category = case_when(
    str_detect(Mixture, "normal") ~ "Healthy",
    str_detect(Mixture, "OA") ~ "OA",
    str_detect(Mixture, "RA") ~ "RA",
    TRUE ~ "Other"
  ))

# Reshape data to long format
long_data <- data %>%
  pivot_longer(cols = 'B cells':'T cells', names_to = "CellType", values_to = "Proportion")

kable(head(long_data))

```
Because of the low T cell population we combine t cells and b cells and call them lymphoids
```{r}
long_data_combined <- long_data %>%
  mutate(CellType = case_when(
    CellType %in% c("B cells", "T cells") ~ "Lymphoids",
    TRUE ~ CellType
  )) %>%
  group_by(Mixture, Category, CellType) %>%
  summarise(Proportion = sum(Proportion), .groups = "drop")
```

Visualize the data with bar plots

For healthy
```{r}
# Stacked bar plot
stacked_bar <- function(df,patient_string) {
  p <- ggplot(df[df$Category == patient_string,], aes(x = Mixture, y = Proportion, fill = CellType)) +
    geom_bar(stat = "identity") +
    scale_fill_brewer(palette = "Set2") +
    labs(
      title = "Cell Type Proportions",
      x = "Tissue Samples",
      y = "Proportion"
    ) +
    theme_minimal() +
    theme(axis.ticks.x=element_blank(),axis.text.x=element_blank()) 
  
  return(p)
}

plot_healthy <- stacked_bar(long_data,"Healthy")
plot_healthy_combined <- stacked_bar(long_data_combined,"Healthy")
plot_healthy_combined
```
For OA
```{r}
plot_oa <- stacked_bar(long_data,"OA")
plot_oa_combined <- stacked_bar(long_data_combined,"OA")

plot_oa_combined
```
For RA
```{r}
plot_ra <- stacked_bar(long_data,"RA")
plot_ra_combined <- stacked_bar(long_data_combined,"RA")
plot_ra_combined
```

Save plots
```{r}
#saving the plot
save_plot <- function(p_h, p_oa, p_ra) {
ggsave(
  plot = p_h,
  filename = "figures/combined/Healthy_top50.png",
  bg = "white",
  width = 7, height = 6
)
ggsave(
  plot = p_oa,
  filename = "figures/combined/OA_top50.png",
  bg = "white",
  width = 7, height = 6
)
ggsave(
  plot = p_ra,
  filename = "figures/combined/RA_top50.png",
  bg = "white",
  width = 12, height = 6
)
}

save_plot(plot_healthy_combined,plot_oa_combined,plot_ra_combined)
```

We can group the cells into three

```{r}
# Calculate the sum of Myeloid Cells and Lymphoids
long_data_groups <- long_data_combined %>%
  group_by(Mixture, Category) %>%
  mutate(ImmuneCellProportion = sum(Proportion[CellType %in% c("Myeloid Cells", "Lymphoids")])) %>%
  ungroup()

long_data_groups <- long_data_groups %>%
  mutate(ImmuneCategory = case_when(
    ImmuneCellProportion < 0.34 ~ "Pauci-immune",
    ImmuneCellProportion >= 0.34 & ImmuneCellProportion <= 0.50 ~ "Moderately\ninfiltrated",
    ImmuneCellProportion > 0.50 ~ "Immune\ncell-dominant",
    TRUE ~ "Other"
  ))

long_data_groups$ImmuneCategory <- factor(long_data_groups$ImmuneCategory, 
                                           levels = c("Pauci-immune", "Moderately\ninfiltrated", "Immune\ncell-dominant"))

write.csv(long_data_groups, "after_deconvolution_bulk_rna.csv")
```

Then visualize them

```{r}
stacked_bar_groups <- function(df,patient_string) {
  
  # Set up the plot with ggplot
  p <- ggplot(df[df$Category == patient_string,], aes(x = Mixture, y = Proportion, fill = CellType)) +
    geom_bar(stat = "identity") +  
    scale_fill_brewer(palette = "Set2") +  
    facet_wrap(~ ImmuneCategory, scales = "free_x") +  # Separate by ImmuneCategory
    labs(
      title = "Cell Type Proportions by Immune Category",
      x = "Tissue Samples",
      y = "Proportion"
    ) +
    theme_minimal() + 
    theme(axis.ticks.x = element_blank(), axis.text.x = element_blank(), panel.grid = element_blank())  
  # Return the plot
  return(p)
}

healthy_groups_plot <- stacked_bar_groups(long_data_groups, "Healthy")
healthy_groups_plot
```
adjusting the widths 

```{r}
  #the numbers from the groups
  c1 <- nrow(long_data_groups[long_data_groups$Category == "Healthy" & long_data_groups$ImmuneCategory == "Pauci-immune",])/4
  c2 <- nrow(long_data_groups[long_data_groups$Category == "Healthy" ,])/4 - c1
  
  #adjust the widths
  healthy_groups_plot_adjusted <- healthy_groups_plot + force_panelsizes(cols = c(c1,c2)) 
  healthy_groups_plot_adjusted
  
  ggsave(
  plot = healthy_groups_plot_adjusted,
  filename = "figures/combined/healthy_grouped.png",
  bg = "white",
  width = 12, height = 6
)

```

same for oa 

```{r}
OA_groups_plot <- stacked_bar_groups(long_data_groups, "OA")

#the numbers from the groups
c1 <- nrow(long_data_groups[long_data_groups$Category == "OA" & long_data_groups$ImmuneCategory == "Pauci-immune",])/4
c2 <- nrow(long_data_groups[long_data_groups$Category == "OA" & long_data_groups$ImmuneCategory == "Moderately\ninfiltrated",])/4
c3 <- nrow(long_data_groups[long_data_groups$Category == "OA" ,])/4 - c1 - c2 
  
# re draw it
OA_groups_plot_adjusted <- OA_groups_plot + force_panelsizes(cols = c(c1,c2,c3)) 
OA_groups_plot_adjusted

  ggsave(
  plot = OA_groups_plot_adjusted,
  filename = "figures/combined/OA_grouped.png",
  bg = "white",
  width = 12, height = 6
)

  
```
finally for ra 

```{r}
RA_groups_plot <- stacked_bar_groups(long_data_groups, "RA")

#the numbers from the groups
c1 <- nrow(long_data_groups[long_data_groups$Category == "RA" & long_data_groups$ImmuneCategory == "Pauci-immune",])/4
c2 <- nrow(long_data_groups[long_data_groups$Category == "RA" & long_data_groups$ImmuneCategory == "Moderately\ninfiltrated",])/4
c3 <- nrow(long_data_groups[long_data_groups$Category == "RA" ,])/4 - c1 - c2 
  
# re draw it
RA_groups_plot_adjusted <- RA_groups_plot + force_panelsizes(cols = c(c1,c2,c3)) 
RA_groups_plot_adjusted

  ggsave(
  plot = RA_groups_plot_adjusted,
  filename = "figures/combined/RA_grouped.png",
  bg = "white",
  width = 16, height = 6
)
```

